Computing solution space properties by generic programming tensor networks \thanks{\funding{...}}

### 2022-03-21 paper/paper.tex
update lattice plot
Updates from Overleaf

\title{Computing properties of independent sets by generic programming tensor networks
\thanks{\funding{...}}
}

@@

\begin{abstract}
%We introduce a method that uses generic programming tensor networks to compute various properties of a class of hard graph problems that can be decribed by local constraints.
We introduce a tensor network algorithm %a method that combines generic programming and tensor network techniques 
to compute various solution space properties of a class of combinatorial optimization problems that can be specified by local constraints. \purple{I still think there is a more concise way to word the class of combinatorial optimization problems this works on...}\blue{changed a bit, better?}
%Let the independent set problem be an example, we show how to compute, for example, the maximum several independent set sizes, the counting of independent sets of a give size, the enumeration of independent sets of a given size.
We look at the independent set problem as an example, and show how to compute the size of the maximum independent set, count the number of independent sets of a given size, and enumerate/sample the independent sets of a given size.
%By making use of generic programming, our algorithms are very simple to implement and one can in addition directly utilize recent advances in tensor network contraction techniques such as near-optimal contraction order finding and slicing to achieve high performance.
By using generic programming techniques, %which require a program to work correctly for different input data types without sacrificing efficiency,
the same simple-to-implement framework can be used to compute all of these properties. Our algorithm utilizes recent advances in tensor network contraction techniques to achieve high performance, including methods to quickly find a near-optimal contraction order and slicing.
The algorithmic complexity of this approach is $2^{{\rm tw}(\overline G)}$, where ${\rm tw} (\overline G)$ is the treewidth of the line graph of the tensor network graph $G$.
Other hard graph problems solvable in our framework include the maximum cut problem, the vertex coloring problem, the maximal clique problem, the dominating set problem, and the satisfiability problem, among others.
To demonstrate how our versatile tool helps to understand these hard problems, we apply it to a few examples, including computing the entropy constant for several hardcore gases on disordered lattices, studying the Overlap Gap Property on unit disk graphs and regular graphs, and visualizing the output of quantum optimization algorithms for the independent set problem. \purple{I might not directly mention the classes of graphs we look at for Overlap Gap - it might be too much information. I'm not sure though, what do you think?} \blue{I agree ``King's graphs at 0.8 filling and 3-regular graphs'' is a bit too detailed, so I rephrased it as ``unit disk graphs and regular graphs'' that only conveys the most important information.} % including the calculations of the entropy constant for some hardcore lattice gases on 2D square lattices and the study of the Overlap Gap Property on three regular graphs.
%We show how our method helps to understand these hard problems on systems with hundreds of vertices.
\end{abstract}

@@

We propose to view the computation of all these properties from an abstract algebra perspective and use the unified framework of
generic tensor-network to solve these seemingly unrelated properties all.
%We map the computation of these properties into generic tensor network contraction with specially designed tensor element algebra.
Tensor network is also known as the sum product network in probabilistic modeling~\cite{Bishop2006} or einsum in linear algebra libraries such as numpy.
It is composed of a set of labelled tensors, where the labels are used for specifying the contraction relation.
The tensor element types are typically restricted to standard number types such as real numbers and complex numbers.
We extend this concept to generic tensor network by generalizing the tensor element type to any type being a commutative semiring.
Given a good contraction order, the space required to contract a tensor network is
two to the power of the treewidth of its line graph~\cite{Markov2008}.
%\purple{I’d write out $2^{{\rm tw}(G)}$}\blue{I feel it is hard to explain line graph without texts.}
Recent progress in simulating quantum circuits with random tensor networks~\cite{Gray2021, Pan2021, Kalachev2021} enables us to find such a good contraction order for a tensor network with thousands of tensors in a reasonable time using heuristic (inexact) methods.
%\purple{Is the scaling of these algorithms known?}\blue{They are not exact algorithms, the time can be any, typically a few minutes, will it be helpful to mention this?}
When using tensor network for solving one of the best solutions,
the time complexity is the same as the dynamic programming algorithm~\cite{Courcelle1990, Fomin2013}.
However, the tensor network has a richer algebraic structure, which crucially enables the computation of solution space properties beyond finding a single solution, which is not possible with dynamic programming.
%These two algorithms are related but are not equivalent.
%Dynamic programming is more general purpose, while a tensor network has a richer algebraic structure, and this structure is crucial for solution space property computation.
%\purple{Not clear how first part and second part of sentence are related}\blue{now? or maybe just removing the comparison with dynamic programming?}
%For sparse graphs, the treewidth is usually much smaller than the number of vertices~\cite{Fomin2006}. Our algorithms, on the other hand, are much more versatile than traditional methods and can be used to compute many other properties than just the MIS size.

@@

\begin{figure}[t] 
    \centering
    \includegraphics[width=\textwidth, trim={0cm 0cm 0cm 0cm}, clip]{figures/lattices.pdf}
    \caption{The types of graphs used in the case study in section \ref{sec:entropy}.
    The lattice dimensions are $L\times L$. (a) square lattice graphs. (b) square lattice graphs with a filling factor $p=0.8$.
    (c) King's graphs. (d) King's graphs with a filling factor $p=0.8$.}
    \label{fig:lattices}
\end{figure}

@@

Which is a clear evidence of that the Overlap Gap Property is absent in $0.8$ filling King's graph.

@@

While in \Fig{fig:hamming}(b), we observed the multiple peak structure as the ratio becomes small for the $3$-regular graphs.

@@

    \caption{The distribution of pairwise Hamming distances for configurations sampled from independent sets with sizes $\geq \lfloor(1-{\rm ratio}) \times \alpha(G)\rfloor$.

@@

>>>>

\title{Computing solution space properties by generic programming tensor networks
\thanks{\funding{...}}
}

@@

\begin{abstract}
%We introduce a method that uses generic programming tensor networks to compute various properties of a class of hard graph problems that can be decribed by local constraints.
We introduce a tensor network algorithm %a method that combines generic programming and tensor network techniques 
to compute various solution space properties of a class of combinatorial optimization problems that can be specified by local constraints. \purple{I still think there is a more concise way to word the class of combinatorial optimization problems this works on...}\blue{changed a bit, better?}
%Let the independent set problem be an example, we show how to compute, for example, the maximum several independent set sizes, the counting of independent sets of a give size, the enumeration of independent sets of a given size.
We look at the independent set problem as an example, and show how to compute the size of the maximum independent set, count the number of independent sets of a given size, and enumerate/sample the independent sets of a given size.
%By making use of generic programming, our algorithms are very simple to implement and one can in addition directly utilize recent advances in tensor network contraction techniques such as near-optimal contraction order finding and slicing to achieve high performance.
By using generic programming techniques, %which require a program to work correctly for different input data types without sacrificing efficiency,
the same simple-to-implement framework can be used to compute all of these properties. Our algorithm utilizes recent advances in tensor network contraction techniques to achieve high performance, including methods to quickly find a near-optimal contraction order and slicing.
%The algorithmic complexity of this approach is two to the treewidth of the line graph of the tensor network graph.
Other hard graph problems solvable in our framework include the maximum cut problem, the vertex coloring problem, the maximal clique problem, the dominating set problem, and the satisfiability problem, among others.
To demonstrate how our versatile tool helps to understand these hard problems, we apply it to a few examples, including computing the entropy constant for several hardcore gases on disordered lattices, studying the Overlap Gap Property on unit disk graphs and regular graphs, and visualizing the output of quantum optimization algorithms for the independent set problem. \purple{I might not directly mention the classes of graphs we look at for Overlap Gap - it might be too much information. I'm not sure though, what do you think?} \blue{I agree ``King's graphs at 0.8 filling and 3-regular graphs'' is a bit too detailed, so I rephrased it as ``unit disk graphs and regular graphs'' that only conveys the most important information.} % including the calculations of the entropy constant for some hardcore lattice gases on 2D square lattices and the study of the Overlap Gap Property on three regular graphs.
%We show how our method helps to understand these hard problems on systems with hundreds of vertices.
\end{abstract}

@@

We propose to view the computation of all these properties from an abstract algebra perspective and use the unified framework of
generic tensor-network to solve these seemingly unrelated properties all.
%We map the computation of these properties into generic tensor network contraction with specially designed tensor element algebra.
Tensor network is also known as the sum product network in probabilistic modeling~\cite{Bishop2006} or einsum in linear algebra libraries such as numpy~\cite{Harris2020}.
It is composed of a set of labelled tensors, where the labels are used for specifying the contraction relation.
The tensor element types are typically restricted to standard number types such as real numbers and complex numbers.
We extend this concept to generic tensor network by generalizing the tensor element type to any type being a commutative semiring.
Given a good contraction order, the space required to contract a tensor network is
two to the power of the treewidth of its line graph~\cite{Markov2008}.
%\purple{I’d write out $2^{{\rm tw}(G)}$}\blue{I feel it is hard to explain line graph without texts.}
Recent progress in simulating quantum circuits with random tensor networks~\cite{Gray2021, Pan2021, Kalachev2021} enables us to find such a good contraction order for a tensor network with thousands of tensors in a reasonable time using heuristic (inexact) methods.
%\purple{Is the scaling of these algorithms known?}\blue{They are not exact algorithms, the time can be any, typically a few minutes, will it be helpful to mention this?}
When using tensor network for solving one of the best solutions,
the time complexity is the same as the dynamic programming algorithm~\cite{Courcelle1990, Fomin2013}.
However, the tensor network has a richer algebraic structure, which crucially enables the computation of solution space properties beyond finding a single solution, which is not possible with dynamic programming.
%These two algorithms are related but are not equivalent.
%Dynamic programming is more general purpose, while a tensor network has a richer algebraic structure, and this structure is crucial for solution space property computation.
%\purple{Not clear how first part and second part of sentence are related}\blue{now? or maybe just removing the comparison with dynamic programming?}
%For sparse graphs, the treewidth is usually much smaller than the number of vertices~\cite{Fomin2006}. Our algorithms, on the other hand, are much more versatile than traditional methods and can be used to compute many other properties than just the MIS size.

@@

\begin{figure}[t] 
    \centering
    \includegraphics[width=\textwidth, trim={0cm 1cm 0cm 0cm}, clip]{figures/lattices.pdf}
    \caption{The types of graphs used in the case study in section \ref{sec:entropy}.
    The lattice dimensions are $L\times L$. (a) square lattice graphs. (b) square lattice graphs with a filling factor $p=0.8$.
    (c) King's graphs. (d) King's graphs with a filling factor $p=0.8$.}
    \label{fig:lattices}
\end{figure}

@@

Which is a clear evidence of the absence of the Overlap Gap Property in $0.8$ filling King's graph.

@@

While in \Fig{fig:hamming}(b), we observed the multiple peak structure in $3$-regular graphs as the ratio becomes small.

@@

    \caption{The plot of pairwise Hamming distances distribution for configurations sampled from independent sets with sizes $\geq \lfloor(1-{\rm ratio}) \times \alpha(G)\rfloor$.

@@

    The ``filled'' region is caused by the fast oscillation of configuration populations of even and odd Hamming distances.
